---
title: "Analyzing RNA-seq Data with the FPKM Package"
author: "Veronica Gosnell"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FPKM Package Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

The `FPKM` package provides utilities for processing and visualizing RNA-seq data using **FPKM** (**F**ragments **P**er **K**ilobase of transcript per **M**illion mapped reads) normalization.

**FPKM** is a commonly used method to normalize RNA-seq data, enabling the comparison of $\underline{gene\space expression}$ levels between samples. It adjusts for both the *length* of the gene and the *total number of reads* in the sample, offering a way to account for **variations** in sequencing depth and gene length. This allows for more accurate comparison of expression levels across genes and samples in RNA-seq experiments.

---

# Introduction

This vignette demonstrates how to use the `FPKM` package to analyze RNA-seq data. We will cover several common tasks, such as reading in RNA-seq data, filtering low-expression genes, normalizing for gene length and sequencing depth, and generating key visualizations such as heatmaps, boxplots, and Principal Component Analysis (PCA).

---

## Installation

First, make sure you have the `FPKM` package installed. You can install it from GitHub with the following command:

```{r installation, results = 'hide', message = FALSE}
devtools::install_github("vsgosnell/FPKM")
```

---

# Loading the Package
## Once installed, load the package and required libraries for this tutorial:

```{r}
library(FPKM)
library(ggplot2)
library(readr)
```

---

# Loading Data

The first step in RNA-seq analysis is to load your data. We will load the RNA-seq data from the `vehicle_drug_feature_counts.txt` file. The file contains read counts for multiple samples, including vehicle (*control*) and drug treatment conditions. 

The `read_rna_seq_data()` function reads in the count matrix and handles any required preprocessing (e.g., removing header metadata).

Download: [vehicle_drug_feature_counts.txt](https://github.com/vsgosnell/FPKM/blob/main/data/vehicle_drug_feature_counts.txt)

```{r read RNAseq data, results = 'hide', message = FALSE}
# Load data from featureCounts output
vehicle_drug_data_path <- "/Users/veronicagosnell/Desktop/R/FPKM/inst/extdata/vehicle_drug_feature_counts.txt"

# Read using readr::read_tsv and skip comment lines
vehicle_drug_data <- readr::read_tsv(vehicle_drug_data_path, comment = "#")

# Rename length column if needed
if ("length" %in% colnames(vehicle_drug_data)) {
  colnames(vehicle_drug_data)[colnames(vehicle_drug_data) == "length"] <- "Length"
}

stopifnot("Length" %in% colnames(vehicle_drug_data))
stopifnot(all(sapply(vehicle_drug_data[, grep("\\.bam$", colnames(vehicle_drug_data))], is.numeric)))
stopifnot(nrow(vehicle_drug_data) == length(vehicle_drug_data$Length))

```

### View the first few rows to confirm it loaded correctly

```{r head RNAseq data}
head(vehicle_drug_data)
```

### This dataset contains the following columns:

- **GeneID**: The gene identifier

- **Chr**: The chromosome

- **Start**: The start position of the gene

- **End**: The end position of the gene

- **Strand**: The positive or negative strand of the gene

- **Length**: The length of the gene

- **vehicle_rep1, vehicle_rep2, drug_rep1, drug_rep2**: 
    - The read counts for different replicates of the vehicle and drug treatments


---

# Calculating FPKM Values

Once the data is loaded, you can compute **FPKM values**. This method is used to normalize gene expression data in RNA-seq experiments by accounting for *sequencing depth* and *gene length*. 

The `calculate_fpkm()` function calculates the **FPKM values** from *read counts* and *gene lengths*.

```{r calculate FPKM, warning = FALSE, message = FALSE}
# Identify sample columns
sample_columns <- grep("\\.bam$", colnames(vehicle_drug_data), value = TRUE)

# Calculate FPKM
fpkm_values <- calculate_fpkm(vehicle_drug_data, count_cols = sample_columns)

# Preview
head(fpkm_values)
```

### Results:

- This function returns a data frame containing the **FPKM values** for each gene across all samples. 

- **FPKM** normalizes expression levels by both gene length (longer genes tend to have more reads) and library size (total read count per sample). This makes expression levels more **comparable** across genes and samples.


---

# Filtering Low Expression Genes

To improve the accuracy of downstream analysis, we can filter out genes with **low expression** levels. 

The `filter_low_expression()` function allows us to filter genes based on a threshold **FPKM** value.


```{r filter low expression}
# Filter out low-expression genes (e.g., average FPKM < 1)
filtered_data <- filter_low_expression(fpkm_values, threshold = 1)
```

### Why filter?

- Genes with consistently **low expression** can introduce *noise*.

- Filtering improves downstream analyses like clustering, PCA, and differential expression.

---

# Visualizing Gene Expression

## Heatmap of Gene Expression

We can create a heatmap to visualize the **gene expression patterns** of the most variable genes across the samples using the `plot_expression_heatmap()` function.


#### **What this does:**

1. **Compute variance** for each gene (row) across your samples

2. **Order** genes by descending variance and pick the top `N`

3. **Subset** your *filtered* FPKM matrix to those genes

4. **Center & scale** each row so differences are on the same scale

5. **Plot** with `pheatmap`, showing just those top `N` genes

$\underline{You\space can\space adjust:}$  
  
  - `N` (e.g. 25, 50, 200)  

  - `show_rownames = TRUE` if you want to see **gene IDs**  

  - `scale = "row"` inside `pheatmap()` instead of *manually* scaling 


```{r heatmap-top-variables, message=FALSE, warning=FALSE, fig.width=8, fig.height=10}
plot_expression_heatmap(filtered_data, top_n = 100)
```

### Results:

  - **Rows** = genes
  
  - **Columns** = samples
  
  - **Color** changes indicate relative *expression differences*

- $\underline{Clustering}$ helps identify patterns of co-expression and sample similarity

---

## Boxplot of FPKM Distribution

A boxplot can be helpful to visualize the distribution of **gene expression** across samples.


```{r FPKM Distribution Boxplot, warning = FALSE, message = FALSE, fig.width=8, fig.height=6}
plot_fpkm_boxplot(filtered_data)
```

### Results:

- Boxplots show **median**, **quartiles**, and **outliers** of log-transformed `FPKM` values

  - Helps assess sample distribution and variability

- RNA-seq data commonly contains lots of genes with 0 or *near-zero* expression

  - These low values create a long lower tail, and the few highly expressed genes pull the median and upper quartile upward

  - So when you plot individual points (like geom_jitter()), they tend to bunch up at the **upper end** of the visible range (i.e., the *top* whisker)
  
  - This can be adjusted according to your needs


---

## Distribution of FPKM Values

To assess the **overall distribution** of expression levels, we can plot the distribution of `FPKM` values.


```{r FPKM Distribution Bar Graph, fig.width=8, fig.height=6}
filtered_data$FPKM <- rowMeans(filtered_data)
plot_fpkm_distribution(filtered_data$FPKM)
```

### Results:

- Most genes have *low* expression (often log-normal distribution)

- Useful for choosing **filtering thresholds** and detecting global shifts

---

# Principal Component Analysis (PCA)

To explore the overall variation in the data, we can perform **Principal Component Analysis (PCA)** on the `FPKM` values. 

**PCA** is used to reduce *dimensionality* and visualize major **patterns** in gene expression data that capture the most important variation across samples. This enables visualization of groupings, patterns, and outliers in a simple 2D plot.


## Quick PCA Visualization

The `FPKM` package provides a convenient wrapper function, `plot_pca_fpkm()`, that performs **log-transformation**, **PCA**, and **plotting** in one step. 

It displays the *first two* principal components, which together explain most of the variance in the data. 

Below, shows both this quick-use function and a step-by-step breakdown of how **PCA** is performed on the `FPKM` dataset.


```{r PCA simple, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
plot_pca_fpkm(filtered_data)
```

### Results:

- Can reveal batch effects, sample outliers, or treatment separation

- Samples that cluster together have **similar expression profiles**

  - If treated and untreated samples cluster separately along PC1 or PC2:

    - This suggests that treatment strongly *influences* gene expression

- If samples from the same group cluster tightly:

  - Indicates *reproducibility* within conditions

- Outliers or overlapping clusters might indicate **batch effects** or **subtle expression changes**


---

## PCA: Step-by-Step Breakdown

While the `plot_pca_fpkm()` function provides a fast and convenient way to visualize **PCA** results, this section walks through the process step-by-step. This approach is useful for users who want to customize the analysis or better understand how **PCA** is computed and interpreted from their `FPKM` data.


```{r PCA breakdown colored, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
# Log-transform filtered data
log_fpkm <- log2(filtered_data + 1)

# transpose the matrix
pca_input <- t(log_fpkm)

# Run PCA
pca_result <- prcomp(pca_input, scale. = TRUE)

# Calculate percentage of variance explained by PC1 and PC2
pve <- round(100 * summary(pca_result)$importance[2, 1:2], 1)

# Create PCA data frame for plotting
pca_df <- as.data.frame(pca_result$x)
pca_df$Sample <- rownames(pca_df)

# Assign conditions based on sample names
pca_df$Condition <- ifelse(grepl("vehicle", pca_df$Sample, ignore.case = TRUE), "Vehicle", ifelse(grepl("drug", pca_df$Sample, ignore.case = TRUE), "Drug", "Unknown"))

# load the ggplot library
library(ggplot2)

# Plot PCA with color by condition
ggplot(pca_df, aes(x = PC1, y = PC2, color = Condition, label = Sample)) +
  geom_point(size = 4) +
  geom_text(vjust = 1.5, size = 4) +
  labs(
    title = "PCA of Gene Expression",
    x = paste0("PC1 (", pve[1], "%)"),
    y = paste0("PC2 (", pve[2], "%)")
  ) +
  theme_minimal()
```

### Results:

  - Implements the full PCA pipeline inline
  
  - Explains and displays PC1/PC2 variance

- PC1 and PC2 capture the **most significant sources of variation** in the dataset

- Samples clustering together on the PCA plot are more similar in expression profiles

- This plot helps identify batch effects, outliers, or biological groupings

### Customizing PCA plot:

The plot can be modified in several ways depending on your needs:

- Add **color** or **shape** by *condition* (i.e., treatment vs control) to visually distinguish biological groups

- Change **scale. argument** to normalize variance across genes

- Use **other Principle Components** (i.e., PC3, PC4, etc.) to analyze more subtle variation

- Add **confidence ellipses** to show group dispersion using stat_ellipse()


---

# Correlation Between Samples

We can also check the correlation between samples using the `plot_sample_correlation()` function.

```{r Correlation Heatmap, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
plot_sample_correlation(filtered_data)
```

### Results:

- Highly correlated samples should cluster *together*

- Useful **quality control** step to confirm experimental replicates behave similarly


---

# Conclusion

This vignette demonstrated how to use the `FPKM` package for RNA-seq data analysis. We successfully:

- Loaded a dataset
- Calculated **FPKM values**
- Filtered low-expression genes
- Visualized the data using:
  - Heatmaps
  - Boxplot
  - PCA

These steps form a robust pipeline for analyzing and interpreting RNA-seq experiments and provide a solid foundation for further exploration of gene expression data.

For advanced usage or more information on the FPKM package, refer to the package documentation and explore additional helper functions.

---
